---
title: "Spatial Interpolation Example"
author: Augustus Pendleton
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Purpose of Document

In our Great Lakes work, we often want to spatially interpolate variables across the lake. Here, I show one approach to doing this, using distance-weighted interpolation from the `gstat` package.

## Installing and Loading Packages and Data

First, we need to install and load packages. Note that some versions **do** matter in this case, as `tmap` is undergoing breaking changes as it transitions to v4. Also, if you're a Mac or Linux user, you may need to separately install the `PROJ`, `GDAL` and/or `GEOS` software. I recommend you do this using homebrew.

```{r install-and-load}
# install.packages("tidyverse")
# install.packages("sf")
# install.packages("terra")
# install.packages("gstat")
# install.packages("remotes")
# install.packages("pacman")
# remotes::install_github("r-tmap/tmap")
# remotes::install_github("mtennekes/tmaptools")

pacman::p_load(tidyverse,
               sf,
               terra,
               gstat,
               tmap,
               tmaptools,
               install = FALSE)
```

Next, we'll load our data objects. First, we have an outline of Lake Ontario, which I downloaded from the USGS. We'll also project it to a Great Lakes specific projection.

```{r load-outline}

ont_outline <- read_sf("data/ontario_outline/hydro_p_LakeOntario.shp")

proj_outline <- ont_outline %>%
                    st_transform(crs = "EPSG:3174") %>%
  st_union()

tm_shape(proj_outline) + 
  tm_borders()

```

Beautiful. Next, we'll load in our data to plot. In this case, it's a csv file which contains temperature measurements for stations across Lake Ontario, collected by the EPA on the September 2023 CSMI cruise.

```{r read-temps}

temp_df <- read_csv("data/temp_measurements.csv")

glimpse(temp_df)

temp_sf <- temp_df %>%
  st_as_sf(coords = c("longitude","latitude"), crs = "EPSG:4326") %>%
  st_transform(crs = "EPSG:3174")

tm_shape(proj_outline) + 
  tm_borders() + 
  tm_shape(temp_sf) + 
  tm_symbols(fill = "temperature")
```

Okay! That's looking good. Let's move on to interpolation.

## Interpolation

Now, we can assume that surface temperature is a continuous variable, and it's generally reasonable to assume that we can interpolate temperatures between stations. This means that our end goal is to go from *vector* data (discrete points) to *raster* data (continuous data defined by the fill of a grid). What we're essentially going to do is use our point measurements to create a function that predicts temperature across a 2D grid. In order to do that, we'll first define a the resolution and spatial extent of the grid, and then map our values across it.

First, let's define our grid.

```{r defining-grid}

# Create an empty raster from our ontario outline
# We'll first need to simplify our outline to an sf, not sfc
outline_sf <- st_sf(proj_outline)

grid <- rast(outline_sf, nrows = 320, ncols = 640)

tm_shape(grid) + 
  tm_raster()
```

Thrilling - we have an empty raster. Changing the nrows and ncols will change the resolution of that grid, which will affect how "smooth" your raster looks at the end. However, higher-resolution grids will take longer to calculate.

In the enxt step, we'll use a spatial model (defined by gstat) to predict the values of a variable at specified locations. Crucially, we can't feed this model a raster grid; rather, it accepts vector points. As such, we need to extract the midpoint of each raster grid cell as a point, and coerce it to an sf object. We'll also filter it such that we only predict values within our Lake Ontario outline.

```{r make_raster_grid}

# Extract mid points
xy <- xyFromCell(grid, 1:ncell(grid))

# Coerce to project point geometry
points_to_predict <- xy %>%
                        as.data.frame(xy) %>%
                        st_as_sf(coords = c("x","y"),
                                 crs = "EPSG:3174") %>%
                        st_filter(proj_outline) # Filter to be within Lake Ontario
```

Now that we've prepped those, we can move to interpolation. First, we build the model, which describes temperature over space, and then we predict the temperature for each of our `points_to_predict`. This is similar to using `lm` to build a linear model, and then using predict to calculate new values in a dataframe.

The crucial tuning parameter here is `idp`, which affects how quickly our temperature prediction decays towards the mean as you travel away from a measure point. This is worth playing around with to achieve a rational-looking map.

```{r interpolation}

# Build the model
ip_model <- gstat(formula = temperature ~ 1, 
                  locations = temp_sf, 
                  nmax = nrow(temp_sf), 
                  set = list(idp = 3))

# Predict temps
temp_preds <- predict(ip_model, points_to_predict) %>% 
  vect() # Needed to work in terra

# Convert points to a raster  
output_raster <- terra::rasterize(temp_preds, grid, field = "var1.pred", fun = "mean")

# Check our new raster
tm_shape(output_raster) + 
  tm_raster()
```

Wow! We just made an interpolated raster of our data. Let's clean up the plotting so we finish with a beautiful map.

```{r final-map}

tm_shape(output_raster) + 
  tm_raster(col.scale = tm_scale_continuous(values = "tol.sunset"),
            col.legend = tm_legend(title = "Temperature (°C)",
                                   orientation = "landscape",
                                   frame = FALSE)) + 
  tm_layout(frame = FALSE) + 
  tm_compass(position = tm_pos_auto_in()) + 
  tm_scalebar(breaks = c(0,25, 50), position = c(.7,0.2), text.size = .6)
```

Looks good! I hope you enjoyed the tutorial.

```{r session-info}
sessionInfo()
```
